---
output: github_document
always_allow_html: yes
editor_options: 
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  message=FALSE,
  warning=FALSE,
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
require(kableExtra)
```
# DasGuptR

The goal of the DasGuptR package is to provide an implementation of Prithwith Das Gupta's specification of *standardization* and *decomposition* of rates[^7].

[^7]: As set out in his 1993 book _[Standardization and decomposition of rates: A user's manual](https://babel.hathitrust.org/cgi/pt?id=osu.32437011198450)_.    
  
## Installation

You can install DasGuptR from here with:

``` r
# install.packages("devtools")
devtools::install_github("josiahpjking/DasGuptR")
```

# Background and worked examples  

Standardization and decomposition are widely used analytic techniques to adjust for the impact of compositional factors on rates.  

* _Standardization_: Shows us what a rate would have been under different scenarios - for example, if there was no difference in the age-structure of the population, or if there was no change in the age-specific rates of the event we are studying.  

* _Decomposition_: Gives us the percentage of the difference in rates between two years attributable to each of the factors we have included in the standardization.  

In the simplest example, consider a case where the rate is taken as the product of two factors $\alpha$ and $\beta$, which we will write as $R = \alpha\beta$. Throughout Das Gupta's work, greek letters are used to indicate the different compositional factors, and upper and lower case latin letters are used to denote specific population values of these:  

|        | pop1 | pop2 |
| ------ | ---- | ---- |
| $\alpha$ | $A$ | $a$ |
| $\beta$ | $B$ | $b$ |
| crude rate | $AB$ | $ab$ |


In this simple case, we can standardise across the two populations, calculating $\alpha$-adjusted rates by replacing $A$ and $a$ with $\frac{a+A}{2}$.  

|        | pop1 | pop2 |
| ------ | ---- | ---- |
| $\alpha$ | $A$ | $a$ |
| $\beta$ | $B$ | $b$ |
| crude rate <br> $R_{crude}$ | $AB$ | $ab$ |
$\beta$-adjusted <br> $R_{-alpha}$ | $A\frac{B+b}{2}$ | $a\frac{B+b}{2}$ | 
$\alpha$-adjusted <br> $R_{-\beta}$ | $\frac{A+a}{2}B$ |  $\frac{A+a}{2}b$ |




These $\alpha$-adjusted rate can be interpreted as "what the crude rate would look like if $\alpha$ was held equal" (and analogously for $\beta$). 
In cases involving multiple factors, this can quickly become unwieldy, as the standardisation process takes each factor in turn and produces an adjusted rate where "all other factors are held equal". For instance, in the case of a rate as the product of four factors $R = \alpha\beta\gamma\delta$, we get out four sets of adjusted rates: the $\alpha\beta\gamma$-adjusted, $\beta\gamma\delta$-adjusted, $\alpha\gamma\delta$-adjusted, and $\alpha\beta\delta$-adjusted.  

For this reason, we opt to refer to these rates as, e.g., $P-\alpha$, where $P$ is the set of all compositional factors. This is reflected in the table above, where we have used $R_{crude}$, $R_{-\alpha}$ and $R_{-\beta}$ to denote the crude, the $P-\alpha$-adjusted (or $\beta$-adjusted), and the $P-\beta$-adjusted (or $\alpha$-adjusted) rates. The $P-\alpha$-adjusted rate can therefore be interpreted as "what the crude rate would look like if $\alpha$ changed but all other factors were held equal". 

The _difference_ in the adjusted (or 'standardised') rates is known as a __decomposition effect__, named due to the fact that differences in the crude rates can be decomposed into differences in adjusted rates: $\Delta R_{crude} = \Delta R_{-\alpha} + \Delta R_{-\beta}$. 
This decomposition allows us to quantify how much of the difference between two crude rates is due to differences in $\alpha$, differences in $\beta$, and so on. [^caus]

[^caus]: Importantly, the use of "attributable" has a very narrow sense of 'numerically attributable', and it is important to stress the lack of any causal interpretation here - different decomposition effects identified by standardization and decomposition may themselves be the products of one (or more) variables not included in the analysis (Das Gupta 1993:4).  


## DasGuptR functionality  

Consider an example of the simple case described above (2 factors, 2 populations). 

DasGuptR requires data to be in long format, with columns for each factor, and a single variable denoting the population:  
```{r}
eg.dg <- data.frame(
  pop = c("pop1","pop2"),
  alpha = c(.6,.3),
  beta = c(.5,.45)
)
eg.dg
```

In this case, the calculations for the adjusted rates can easily be calculated manually: 

```{r}
data.frame(
  pop = c("pop1","pop2"),
  Rcrude = c(.6*.5, .3*.55),
  R_alpha = c(.6,.3) * ((.5+.45)/2),
  R_beta = ((.6+.3)/2) * c(.5,.45)
)
```

The workhorse of the DasGuptR package is `dgnpop()` which computes the adjusted rates for P factors across N populations:  

```{r}
library(DasGuptR)
dgnpop(eg.dg, pop = "pop", factors=c("alpha","beta"))
```

These can be quickly turned into a wide table in the style of Das Gupta using `dg_table()`. 
When working with just two populations, this also provides the difference in rates, also expressed as the percentage of the crude rate difference:  

```{r}
dgnpop(eg.dg, pop = "pop", factors=c("alpha","beta")) |>
  dg_table()
```

`dg_plot()` will plot these (although this is really only useful when working with many populations, such as with time series data)
```{r}
dgnpop(eg.dg, pop = "pop", factors=c("alpha","beta")) |>
  dg_plot()
```
  
  

## Rates as functions, N populations, vector factors and cross-classified data
  
Rates may be composed of many factors, and they may not be calculated as a simple product. Additionally, we may desire to standardise across many populations --- such as in a time series --- or we may have be interested in how the compositional structure of populations contributes to the differences in these rates.

The full explanation of Das Gupta's methodology for standardisation and decomposition are explained in full in his 1993 book _[Standardization and decomposition of rates: A user's manual](https://babel.hathitrust.org/cgi/pt?id=osu.32437011198450)_ 

Below are various examples taken from Das Gupta's 1993 work, following his exposition by building up the number of factors, vectorisation, generalising to different rate functions, compositional structures of populations, and finally extending the process to more than just two populations.  


### P factors

The addition of factors into the make up of a population rate is handled in `dgnpop()` by simply adding the variable name into the `factors` argument.  
The default behaviour will take the rate to be the product of all factors specified.  

#### 2 factors, 2 populations, R=ab

```{r}
eg2.1 <- data.frame(
  pop = c("black","white"),
  avg_earnings = c(10930, 16591),
  earner_prop = c(.717892, .825974)
)

dgnpop(eg2.1, pop="pop", factors=c("avg_earnings", "earner_prop")) |>
  dg_table()
```

#### 3 factors, 2 populations, R=abc

```{r}
eg2.3 <- data.frame(
  pop = c("austria","chile"),
  birthsw1549 = c(51.78746, 84.90502),
  propw1549 = c(.45919, .75756),
  propw = c(.52638, .51065)
)

dgnpop(eg2.3, pop="pop", factors=c("birthsw1549", "propw1549", "propw")) |>
  dg_table()
```


#### 4 factors, 2 populations, R=abcd

```{r}
eg2.5 <- data.frame(
  pop = c(1971, 1979),
  birth_preg = c(25.3, 32.7),
  preg_actw = c(.214, .290),
  actw_prop = c(.279, .473),
  w_prop = c(.949, .986)
)

dgnpop(eg2.5, pop="pop", c("birth_preg", "preg_actw", "actw_prop", "w_prop")) |>
  dg_table()
```

#### 5 factors, 2 populations, R=abcde

```{r}
eg2.7 <- data.frame(
  pop = c(1970, 1980),
  prop_m = c(.58, .72),
  noncontr = c(.76, .97),
  abort = c(.84, .97),
  lact = c(.66, .56),
  fecund = c(16.573, 16.158)
)
dgnpop(eg2.7, pop="pop", 
       factors = c("prop_m", "noncontr", "abort", "lact", "fecund")) |>
  dg_table()
```

### vector factors   

Often, we may have data for each compositional factor on a set of sub-populations, and the crude rates for the population are the aggregated cell-specific rates.  
In these cases, `dgnpop()` requires the user to provide both the variable(s) indicating the sub-populations, and an appropriate rate function that aggregates up to a summary value for each population.  
For instance, in the example below, the cell-specific rates are calculated as the product of 3 factors, and the population rate is the sum of the cell-specific rates, so the user would specify `ratefunction="sum(a*b*c)"`. 

#### 3 vector factors, 2 populations, R=sum(abc)

```{r}
eg4.5 <- data.frame(
  agegroup = rep(1:7, 2),
  pop = rep(c(1970, 1960), e = 7),
  bm = c(488, 452, 338, 156, 63, 22, 3,
         393, 407, 369, 274, 184, 90, 16),
  mw = c(.082, .527, .866, .941, .942, .923, .876,
         .122, .622, .903, .930, .916, .873, .800),
  wp = c(.058, .038, .032, .030, .026, .023, .019,
         .043, .041, .036, .032, .026, .020, .018)
)
dgnpop(eg4.5, pop="pop", c("bm", "mw", "wp"),
       ratefunction = "sum(bm*mw*wp)") |>
  dg_table()
```

Note that for most purposes, working with vector factors would require the rate function to aggregate up to a summary, population-level value, e.g., using `"sum(a*b*c)"`. If this is not provided, then `dgnpop()` will return an array of standardised sub-population rates, of the same length as the number of sub-populations. In order to do this, the user is also required to specify the variable indicating the sub-population in `id_vars` argument.  
```{r}
dgnpop(eg4.5, pop="pop", c("bm", "mw", "wp"), id_vars=c("agegroup"),
       ratefunction = "bm*mw*wp")
```

In this case, we can retrieve the population rates by summing up:  
```{r}
library(tidyverse)

dgnpop(eg4.5, pop="pop", c("bm", "mw", "wp"), id_vars=c("agegroup"),
       ratefunction = "bm*mw*wp") |>
  group_by(pop,factor) |>
  summarise(
    rate = sum(rate)
  )
```

Equivalently, in this simple case where we have just one variable indicating a single set of sub-populations (e.g., different age groups), then we could run `dgnpop()` on each sub-population individually, and aggregate up post-hoc:  
```{r}
eg4.5 |> 
  nest(data=-agegroup) |>
  mutate(
    dg = map(data, ~ dgnpop(., pop="pop",factors=c("bm","mw","wp")))
  ) |> 
  unnest(dg) |>
  group_by(pop,factor) |>
  summarise(
    rate = sum(rate)
  ) |> dg_table()
```


### rates as functions

The `ratefunction` argument of `dgnpop()` essentially allows the user to define a custom rate function. This may be as simple as $R = \alpha - \beta$ (specified as `ratefunction = "a-b"`):  

```{r}
eg3.1 <- data.frame(
  pop = c(1940,1960),
  crude_birth = c(19.4, 23.7),
  crude_death = c(10.8, 9.5)
)
dgnpop(eg3.1, pop="pop",c("crude_birth","crude_death"),
       ratefunction = "crude_birth-crude_death") |>
  dg_table()
```

However, it may be something more complex. For instance, when working with vector factors, we might define the rate as in various ways that include aggregating over sub-populations in different combinations of factors (e.g., `sum(a*b)/sum(a*b*c)`).  

The example below shows once such example:  
```{r}
# 4 vector factors, 2 populations, R=f(A,B)
# rate as function of vector factors
eg4.4 <- data.frame(
  pop=rep(c(1963,1983),e=6),
  agegroup=c("15-19","20-24","25-29","30-34","35-39","40-44"),
  A = c(.200,.163,.146,.154,.168,.169,
        .169,.195,.190,.174,.150,.122),
  B = c(.866,.325,.119,.099,.099,.121,
        .931,.563,.311,.216,.199,.191),
  C = c(.007,.021,.023,.015,.008,.002,
        .018,.026,.023,.016,.008,.002),
  D = c(.454,.326,.195,.107,.051,.015,
        .380,.201,.149,.079,.025,.006)
)
dgnpop(eg4.4, pop="pop",factors=c("A","B","C","D"), id_vars = "agegroup",
       ratefunction="sum(A*B*C) / (sum(A*B*C) + sum(A*(1-B)*D))") |>
  dg_table()
```

The `ratefunction` argument can be given any string that, when parsed and evaluated, will return a summary value for a rate. At the point at which the string is evaluated, each factor (or vector-factor) is stored in a named list, meaning the function must simply refer to those factors by name.  

It is possible, for instance, to define a custom function in the user's environment, and provide a call to that function to the `ratefunction` argument of `dgnpop()`:  
```{r}
myratef <- function(a,b,c,d){
  return( sum(a*b*c) / (sum(a*b*c) + sum(a*(1-b)*d))  )
}

dgnpop(eg4.4, pop="pop",factors=c("A","B","C","D"), id_vars = "agegroup",
       ratefunction="myratef(A,B,C,D)") |>
  dg_table()
```

The upshot of this is that there is not really any limit of the complexity to the rate function the user wishes to specify. Das Gupta provides one such example in which the rate is obtained iteratively via Newton-Raphson:  

```{r}
eg4.1 <- data.frame(
  age_group = c("10-15","15-20","20-25","25-30","30-35","35-40","40-45","45-50","50-55"),
  pop=rep(c(1965,1960),e=9),
  Lx=c(486446,485454,483929,482046,479522,475844,470419,462351,450468,
       485434,484410,492905,481001,478485,474911,469528,461368,449349),
  mx=c(.00041,.03416,.09584,.07915,.04651,.02283,.00631,.00038,.00000,
       .00040,.04335,.12581,.09641,.05504,.02760,.00756,.00045,.00000)
)

# rate function:  
RF4.1 <- function(A,B){
  idx = 1:length(A)
  mu0 = sum(A*B/100000)
  mu1 = sum((5*idx + 7.5)*A*B/100000)
  r1 = log(mu0) * (mu0/mu1)
  while(TRUE){
    Nr1 = 0
    Dr1 = 0
    Nr1 = Nr1 + sum(exp(-r1 * (5*idx + 7.5)) * A * (B / 100000))
    Dr1 = Dr1 - sum((5*idx + 7.5) * exp(-r1 * (5*idx + 7.5)) * A * (B / 100000))
    r2 = r1 - ((Nr1 - 1)/Dr1)
    if(abs(r2 - r1)<=.0000001){
      break
    }
    r1 = r2
  }
  return(r2)
}
# crude rates:
RF4.1(A = eg4.1$Lx[1:9], B = eg4.1$mx[1:9])
RF4.1(A = eg4.1$Lx[10:18], B = eg4.1$mx[10:18])

# decomposition: 
dgnpop(eg4.1, pop="pop",factors=c("Lx","mx"), id_vars = "age_group",
       ratefunction="RF4.1(Lx,mx)") |>
  dg_table()
```

### population structures and cross-classified data  

Very often, the splitting up of a population into various sub-populations is done because we are interested in separating out how much the population rate differences are due to difference in the compositional structure of the population vs differences in the cell-specific rates.  

To do this, we require data on the sizes (or relative sizes) of each sub-population. The simplest case here would be to disagreggate into a single set of sub-populations (e.g., age-groups), and simply have cell-specific rates themselves. The crude rates for the population are therefore simply the sum of all the cell-specific rates weighted by the relative size of the cell.  

In the example below, the relative size of each cell is provided as a percentage in the `size` column.  
```{r}
eg5.1 <- data.frame(
  age_group = rep(c("15-19","20-24","25-29","30-34","35-39",
                "40-44","45-49","50-54","55-59","60-64",
                "65-69","70-74","75+"),2),
  pop = rep(c(1970,1985),e=13),
  size = c(12.9,10.9,9.5,8.0,7.8,8.4,8.6,7.8,7.0,5.9,4.7,3.6,4.9,
           10.1,11.2,11.6,10.9,9.4,7.7,6.3,6.0,6.3,5.9,5.1,4.0,5.5),
  rate = c(1.9,25.8,45.7,49.6,51.2,51.6,51.8,54.9,58.7,60.4,62.8,66.6,66.8,
           2.2,24.3,45.8,52.5,56.1,55.6,56.0,57.4,57.2,61.2,63.9,68.6,72.2)
)
```

We can decompose this into the rate-adjusted and age-adjusted rates in various ways. 

1. Creating a new column of proportions (rather than percentages) we can include it in the list of factors and do decomposition as previously:  

```{r}
eg5.1$age_str <- eg5.1$size/100

dgnpop(eg5.1, pop="pop",factors=c("age_str","rate"), id_vars = "age_group",
       ratefunction="sum(age_str*rate)") |>
  dg_table() 
```

2. We could alternatively simply include the conversion to proportion in the rate function. The inputs to the rate function here are the set of vector factors for each population, so internal calls to `sum(size)` will give us the total population size:  
```{r}
dgnpop(eg5.1, pop="pop",factors=c("size","rate"), id_vars = "age_group",
       ratefunction="sum( (size/sum(size))*rate )") |>
  dg_table()
```

3. Finally, we can instead provide the variable indicating the size of each sub-population into the `crossclassified` argument of `dgnpop()`. 
```{r}
dgnpop(eg5.1, pop="pop", factors=c("rate"), id_vars = "age_group",
       crossclassified = "size") |>
  dg_table() 
```


This latter approach can be extended to situations in which we have cross-classified data - i.e. individual sub-populations are defined by the combination of multiple variables such as age and race.  

```{r}
eg5.3 <- data.frame(
  race = rep(rep(1:2, e=11),2),
  age = rep(rep(1:11,2),2),
  pop = rep(c(1985,1970), e=22),
  size = c(3041,11577,27450,32711,35480,27411,19555,19795,15254,8022,2472,
           707,2692,6473,6841,6547,4352,3034,2540,1749,804,236,
           2968,11484,34614,30992,21983,20314,20928,16897,11339,5720,1315,
           535,2162,6120,4781,3096,2718,2363,1767,1149,448,117),
  rate = c(9.163,0.462,0.248,0.929,1.084,1.810,4.715,12.187,27.728,64.068,157.570,
           17.208,0.738,0.328,1.103,2.045,3.724,8.052,17.812,34.128,68.276,125.161,
           18.469,0.751,0.391,1.146,1.287,2.672,6.636,15.691,34.723,79.763,176.837,
           36.993,1.352,0.541,2.040,3.523,6.746,12.967,24.471,45.091,74.902,123.205)
)
```

In these cases, using the sub-population relative sizes as a compositional factor straight off the bat will collapse the two variables that make up the structure of the population: 
```{r}
dgnpop(eg5.3, pop = "pop", factors=c("size","rate"), id_vars = c("race","age"),
       ratefunction = "sum( (size/sum(size))*rate)") |>
  dg_table() 
```

Instead, providing the cell-specific sizes to the `crossclassified` argument will re-express the proportion of the population in a given cell as a product of K factors representing each of the structural variables. These are then included in the decomposition process:  

```{r}
dgnpop(eg5.3, pop = "pop", factors=c("rate"), id_vars = c("race","age"),
       crossclassified = "size") |>
  dg_table() 
```





::: {.callout-tip collapse="true"}
#### re-expression of population structures:

```{r}
library(tidyverse)
eg5.3a <- eg5.3 |>
  group_by(pop) |> mutate(n_tot = sum(size)) |>
  group_by(pop,age) |> mutate(n_age = sum(size)) |>
  group_by(pop,race) |> mutate(n_race = sum(size)) |>
  ungroup() |>
  mutate(
    A = ((size / n_race) * (n_age / n_tot))^(1/2),
    B = ((size / n_age) * (n_race / n_tot))^(1/2),
    AB = A*B, # same as:
    prop = size/n_tot
  )
head(eg5.3a)

dgcc(eg5.3, "pop", id_vars=c("age","race"), crossclassified ="size") |>
  head()

dgnpop(eg5.3a, pop = "pop", factors=c("A","B","rate"), id_vars = c("race","age"),
       ratefunction = "sum( (A*B*rate)/sum(A*B) )") |>
  dg_table()
```

:::

### N populations

```{r}
# 4 vector factors, 5 populations, R=f(A,B)
# rate as function of vector factors
eg6.6 <- data.frame(
  pop=rep(c(1963,1968,1973,1978,1983),e=6),
  agegroup=c("15-19","20-24","25-29","30-34","35-39","40-44"),
  A = c(.200,.163,.146,.154,.168,.169,
        .215,.191,.156,.137,.144,.157,
        .218,.203,.175,.144,.127,.133,
        .205,.200,.181,.162,.134,.118,
        .169,.195,.190,.174,.150,.122),
  B = c(.866,.325,.119,.099,.099,.121,
        .891,.373,.124,.100,.107,.127,
        .870,.396,.158,.125,.113,.129,
        .900,.484,.243,.176,.155,.168,
        .931,.563,.311,.216,.199,.191),
  C = c(.007,.021,.023,.015,.008,.002,
        .010,.023,.023,.015,.008,.002,
        .011,.016,.017,.011,.006,.002,
        .014,.019,.015,.010,.005,.001,
        .018,.026,.023,.016,.008,.002),
  D = c(.454,.326,.195,.107,.051,.015,
        .433,.249,.159,.079,.037,.011,
        .314,.181,.133,.063,.023,.006,
        .313,.191,.143,.069,.021,.004,
        .380,.201,.149,.079,.025,.006)
)
# crude rates
eg6.6 |> group_by(pop) |>
  summarise(
    crude = sum(A*B*C) / (sum(A*B*C) + sum(A*(1-B)*D))
  )

# decomposition
dgnpop(eg6.6, pop="pop",factors=c("A","B","C","D"),id_vars="agegroup",
       ratefunction="1000*sum(A*B*C) / (sum(A*B*C) + sum(A*(1-B)*D))")$rates |>
  dg_plot()


dgnpop(eg6.6, pop="pop",factors=c("A","B","C","D"),id_vars="agegroup",
       ratefunction="1000*sum(A*B*C) / (sum(A*B*C) + sum(A*(1-B)*D))")$rates |>
  dg_table()

dgnpop(eg6.6, pop="pop",factors=c("A","B","C","D"),id_vars="agegroup",
       ratefunction="1000*sum(A*B*C) / (sum(A*B*C) + sum(A*(1-B)*D))")$rates |>
  dg_table(pop1="1963",pop2="1968")

dgnpop(eg6.6, pop="pop",factors=c("A","B","C","D"),id_vars="agegroup",
       ratefunction="1000*sum(A*B*C) / (sum(A*B*C) + sum(A*(1-B)*D))")$diffs |>
  select(-pop,-std.set) |>
  pivot_wider(values_from=diff,names_from=diff.calc)
```


## US population data

```{r}

ratedata <- read.csv("archive/uspop_dg.csv")
dgous <- dgnpop(ratedata, pop="year",factors=c("birthrate"),id_vars="agebin",
                crossclassified="thous",ratefunction="birthrate")
dg_plot(dgous$rates)


```

## scottish reconviction data

```{r}
data(reconv)
str(reconv)

dg_rec <- dgnpop(reconv, pop="year", 
                 factors=c("prev_rate"),id_vars=c("Sex","Age"),
                 crossclassified="offenders")
dg_plot(dg_rec$rates)


```



